Penalized regression for left-truncated and right-censored survival data

### 2021-06-24 manuscript.tex
Minor edits to text

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \cite{tibshirani1996regression} and $l_2$ (ridge) \cite{tikhonov1963ridge} regression, elastic net regression \cite{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \cite{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors. In a high-dimensional feature space, where the number of predictors $p$ is large relative to the sample size $n$ , a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting. Note that in high-dimensional settings where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a standard regression cannot be fit at all.

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with LTRC data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., excluded from the sample) if $V_i > Y_i$.

@@

Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e., non-truncated) patients. Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected. Models were fit with and without adjusting for left truncation. All models adjusted for right censoring.

@@

The smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR). Short variants (SV) were considered for each biomarker while copy number alterations (CN) and rearrangements (RE) were also considered for TP53. $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.

@@

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.

@@

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models tended to push the most hazard ratios toward $1$, resulting in less separation of low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.

@@

Our results have important implications for specification, interpretation, and evaluation of prognostic survival models. Much of the difficulty arises because training and test sets are both subject to the same selection bias. In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray. For example, our simulations suggest that the commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of LTRC data. On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.

In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While some of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

@@

When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \cite{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \cite{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \cite{thiebaut2004choice, chiou2019transformation}.

>>>>

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \cite{tibshirani1996regression} and $l_2$ (ridge) \cite{tikhonov1963ridge} regression, elastic net regression \cite{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \cite{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors. In a high-dimensional feature space, where the number of predictors $p$ is large relative to the sample size $n$ , a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting. Note that in high-dimensional settings where the number of predictors is larger than the sample size (i.e. $p$ > $n$), a standard regression cannot be fit at all.

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with LTRC data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e. excluded from the sample) if $V_i > Y_i$.

@@

Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e. non-truncated) patients. Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected. Models were fit with and without adjusting for left truncation. All models adjusted for right censoring.

@@

The smaller model consisted of the following variables: year of diagnosis (i.e. index year), age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR). Short variants (SV) were considered for each biomarker while copy number alterations (CN) and rearrangements (RE) were also considered for TP53. $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.

@@

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e. to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is reduced to $1$ or less than $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.

@@

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models tended to push hazard ratios---especially for the influential ``year of diagnosis'' predictor---toward $1$, resulting in less separation of low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.

@@

Our results have important implications for specification, interpretation, and evaluation of prognostic survival models. Much of the difficulty arises because training and test sets are both subject to the same selection bias. In other words, a careful understanding of the data generating process is crucial and naive approaches can lead analysts astray. For example, our simulations suggest that the commonly used C-index cannot differentiate between models that mimic the true data generating process and those that do not, and may therefore lead to misleading conclusions in the presence of LTRC data. On the other hand, metrics that focus on predicted survival probabilities such as calibration curves are better able to determine whether a model can accurately predict the risk of an event.  
In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While some of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

@@

When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e. when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \cite{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \cite{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \cite{thiebaut2004choice, chiou2019transformation}.
